---
title: Working with data in R! 
format: html
---
## Introduction

In the last exercise, we learned that the purpose of a Markdown file is to keep explanations for computational problems and the code we use to solve them next to each other in the same document. 

We had a couple 'chunks' of text, code, text, code, etc, but we didn't have anything very interesting to look at. 

If we go a little farther with the tools available in R, and grab a dataset, R might make itself more worthwhile! 


##Let's view a dataset 
There's a really handy sample dataset about diamonds that comes along with one of R's most useful plotting tools!

The tool is called ggplot, so I'll show you how to get the toolset and the dataset with a simple command. 

```{r ExploreDIAMONDS}

#To read some pieces of code that other people have already written, we can borrow those codebooks from the R library. 

#Here, we borrow some instructions for making plots, which also includes the sample dataset called 'diamonds', by calling up the code using the 'library' function:  
library(ggplot2)

#This is like checking out a codebook from the library - you get to keep it at hand as long as you're working in this space.

#The 'head' command in bash has the same name in R, but it is a function too: 
head(diamonds)

#To change the number of rows displayed, it looks like this: 
head(diamonds, n=10)

#We also talked about how the function nrow() works just like the bash command (word count) that tells us how many lines are in a file: wc -l
nrow(diamonds)
```


##Let's navigate a dataset 
In lecture we saw how you could find things in a dataframe by specifying rows and columns with this type of syntax: 

dataframe[rows, columns]

In addition, you can put functions inside the row and column spots. The function we used as an example is the 'combine' function: c()

dataframe[c(1,3,5,7,9), ] returns the odd rows less than 10 from the dataframe 

In fact, even grep is available in R as a function so you can search for patterns inside your data. Typically, you want to return the rows that match, because these are complete records of whatever thing your studying. 

What does the code look like using the real data about the R  diamonds? Let's investigate: 

```{r DiaDataFrames}

#This is an example of some student code where they were looking for all the "Good" diamonds: pretty close! 
diamonds[grep("Good",diamonds)]

#Putting results in a variable is good practice if we want to us them later. 

# Use the '<-' combined symbol to "stuff" your code result from the right side into the variable on the left side. You can name the variable whatever you want! We named it 'stud' here: 
stud <- diamonds[grep("Good",diamonds)]

#Let us make a few modifications. 
#First, let's specify that we are looking in rows with a comma after the grep command
#Then let's grep for the pattern in the "cut" column only. 

#To use column information inside the function, we can use SYNTAX like this:
#diamonds$cut. That calls the column "by its name". 

stud2 <- diamonds[grep("Good",diamonds$cut),]
stud2

#Here is a student example of "subsetting" a dataset. 
#This can be a good idea to figure out how the commands work by using smaller datasets.
diamonds[1:10, c("cut","price")] -> dia
dia

#The student code works, but the more typical way to write this is in R is to put the variable name first and the instructions second: 

dia2 <- diamonds[1:10, c("cut", "price")]
dia2


#Much like the 'MATCH' function in excel, the grep function in R returns the index of the matches. 
#Consider this grep function, which searches for the word 'Premium' in the dataframe 'dia2'. Where does it say the matches are? 
matches <- grep("Premium", dia2$cut)
matches

#You should see all of the row numbers returned. So how do we get the information in the rows returned, instead of just which rows have a match? 

#We might try to just use the returned indexes to grab rows: 
dia2[matches,]

#That works! 

#You could also try using the combine function we saw earlier, because we got more than one row number back: 
dia2[c(matches),]

#That works too!

#Finally, let's see if we can stuff the function inside the dataframe to select the right rows, alll in one line of code: 
dia2[grep("Premium", dia2$cut), ]

#I hope you can see that there are LOTS of ways to do things 'right', not just LOTS of ways to get it a little wrong. 

#Another version of grep that students have tried is 'grepl', which returns something called a boolean. 

#Booleans are either 'TRUE' or 'FALSE' for each of the data items surveyed. 

#This could be useful, let's save that list of values in a variable and try some of the same steps as above. 

stud3 <- grepl("Premium", dia2$cut)
stud3
dia2[stud3,]

#Ah! Both booleans from grepl and indexes from grep work!

#Now, I told you the fun part was plotting. How do we? Try this code: 

ggplot(data = dia, aes(x = cut, y = price, color = cut)) +
  geom_point(alpha = 0.8, size =4)

#You'll notice we need to call out plotting function ggplot(). 

#The function needs a dataset (dia), and the variables to use get packaged into the 'aes()' function (aes = aesthetics; how modern!).

#The handy part of ggplot is that we can add-on instructions with extra functions using the '+' sign - kinda like buying upgrades, but free! 

#Here, we add-on upgrades to our datapoints using the function 'geom_point()' and give it a new color opacity using 'alpha' and a new point size using 'size'. Try to change some values and see what happens!  

#For the TASKS, I've created a subset of diamonds called diatask with this command: 

diatask <- diamonds[80:100, c(2,7)]
#print(diatask, n = 20) 

```

## Tasks 

For the following tasks, put your answers in the code chunk below each question. 

1. What are the key variables in diaTask? Hint: Use the head function and then write the key variables as comments below your code. 

```{r question1}

```

2. Use a combination of nrow() and head() to display ALL of the rows of the diatask dataframe.  

```{r question2}

```

3.Use a mathematical formula to display all of the rows of diatask where the diamond price is less than $750. 


```{r question3}

```

4. In R, there is a function that does the same thing as the '>>' redirection in bash, i.e. it APPENDS data. The R function is called rbind(). 

Use your knowledge of grep in R, and the function rbind(), to create a NEW dataframe called 'sparkles' from diatask that has BOTH 'Premium' and 'Ideal' diamonds in it, but no other cuts. 

A. Start your codechunk with a comment on how many rows 'sparkles' *should* have, by looking at it! 

HINT: You may not be "seeing" all the rows by default. I left a piece of code for you in the comments that uses the 'print' function to show them. Go turn it from a comment into code to use it! This is called 'commenting out' code, and is a useful trick to hide things you're working on from the computer. 

B. Now write your code to display the dataframe with the correct records.  

C. Below your code, write a comment to tell me if you got the same number of records (rows) that you expected from part A.  

```{r question4}

```

5. Use ggplot to display the information from the diatask dataframe.

HINT: Copy and paste the code, and then just change the piece you need to update! 


```{r question5}

```

6. How do you remove the ugly grey gridlines from the ggplot? 

HINT: Look up a function that you can add-on to make this easier. 

```{r question6}

```

7. Change the diatask command so that it contains the carat of the diamonds too, and name it diasize. 


```{r question7}

```

8. Show the plot of your new dataset, diasize, with carat on the x-axis and price on the y-axis. 

HINT: Copy and paste the code, and then just change the piece you need to update! 

```{r question8}

```

9. Why are the plots for diatask and diasize insufficient to make a conclusion about the relationships between cut and price or carat and price? 

```{r question9}
## Write your answer as a comment here
```

10. Create your own plots for either cut OR carat on the x-axis, with price still on the y-axis. Make sure to construct the dataset so that you can make a strong conclusion about how your chosen independent variable affects price. 

Write a response in the comments: what did you do to construct the dataset and were you surprised by the results? 

HINT: Copy and paste the code, and then just change the piece you need to update! 

```{r question10}


```


## Notes on grep attempts: 

1) The pattern you specify must EXACTLY match the value in the dataset, unless you have allowed for 'fuzziness'.
  
  Examples: 
    grep("ideal", diamonds) is NOT the same as grep("Ideal", diamonds)
    Spelling counts! Misspelled words will not be recognized in either the commands or the dataset
  
2) Make sure you use proper SYNTAX 
  
  Examples: 
  dataframe[,]
  function()
  variable <- dataframe[1:100,]

3) You can use e.g. column names to tell grep where to look for the pattern: 

  grep("Good", diamonds$cut)