---
title: "Working with Data in R!"
format: html
---

## Introduction

In the last exercise, we learned that the purpose of a Markdown file is to keep explanations for computational problems and the code we use to solve them next to each other in the same document.

We had a couple of chunks of text, code, text, and code, but we did not have anything very interesting to look at.

If we go a little farther with the tools available in R and use a dataset, R can become much more useful.

## Let's view a dataset

There is a handy sample dataset about diamonds that comes with one of R's most useful plotting packages.

The package is called **ggplot2**, so the following code loads the package and its `diamonds` dataset.

```{r ExploreDIAMONDS}
# Load ggplot2, which includes the example dataset named diamonds.
library(ggplot2)

# Display the first six rows.
head(diamonds)

# Display the first ten rows.
head(diamonds, n = 10)

# Count the number of rows in the complete dataset.
nrow(diamonds)
```

## Let's navigate a dataset

In lecture, we saw how to find values in a data frame by specifying rows and columns with this syntax:

```text
dataframe[rows, columns]
```

You can also put functions inside the row and column positions. The function used in the example is the combine function, `c()`.

```text
dataframe[c(1, 3, 5, 7, 9), ]
```

This returns the odd-numbered rows below 10 and keeps every column.

The `grep()` function can search for patterns inside data. Usually, we want to return complete rows that match the pattern.

```{r DiaDataFrames}
# These two lines are intentionally commented out because they do not
# specify the column that grep() should search or the data-frame dimension
# that should be returned. Leaving them active would stop the document.
#
# diamonds[grep("Good", diamonds)]
# stud <- diamonds[grep("Good", diamonds)]

# Search the cut column and use the returned row positions to subset diamonds.
stud2 <- diamonds[grep("^Good$", diamonds$cut), ]
stud2

# Create a smaller data frame containing rows 1 through 10
# and only the cut and price columns.
dia <- diamonds[1:10, c("cut", "price")]
dia

# The more common R assignment style places the object name on the left.
dia2 <- diamonds[1:10, c("cut", "price")]
dia2

# grep() returns the row positions where "Premium" occurs.
matches <- grep("^Premium$", dia2$cut)
matches

# Use those positions to retrieve complete rows.
dia2[matches, ]

# c(matches) also works, although c() is unnecessary because
# matches is already a vector.
dia2[c(matches), ]

# Put grep() directly inside the row position.
dia2[grep("^Premium$", dia2$cut), ]

# grepl() returns TRUE or FALSE for each value.
stud3 <- grepl("^Premium$", dia2$cut)
stud3
dia2[stud3, ]

# Plot the smaller data frame.
ggplot(data = dia, aes(x = cut, y = price, color = cut)) +
  geom_point(alpha = 0.8, size = 4)

# Create the data frame used for Tasks 1 through 6.
# R uses one-based indexing, so this contains rows 80 through 100.
diatask <- diamonds[80:100, c(2, 7)]

# Display every row so the records can be inspected.
print(diatask, n = nrow(diatask))
```

## Tasks

### 1. Identify the key variables in `diatask`

```{r question1}
head(diatask)

# The key variables are cut and price.
# cut describes the quality category of the diamond's cut.
# price gives the diamond's price in US dollars.
```

### 2. Display all rows of `diatask`

```{r question2}
head(diatask, n = nrow(diatask))
```

### 3. Display rows where price is below $750

```{r question3}
diatask[diatask$price < 750, ]
```

### 4. Create `sparkles` with only Premium and Ideal diamonds

```{r question4}
# By inspecting diatask, sparkles should contain 10 rows:
# 6 Premium diamonds and 4 Ideal diamonds.

premium_diamonds <- diatask[grep("^Premium$", diatask$cut), ]
ideal_diamonds <- diatask[grep("^Ideal$", diatask$cut), ]

sparkles <- rbind(premium_diamonds, ideal_diamonds)

sparkles
nrow(sparkles)

# Yes. The result contains the 10 rows that I expected.
```

### 5. Plot the information in `diatask`

```{r question5}
ggplot(data = diatask, aes(x = cut, y = price, color = cut)) +
  geom_point(alpha = 0.8, size = 4) +
  labs(
    title = "Diamond Price by Cut",
    x = "Cut",
    y = "Price (US dollars)"
  )
```

### 6. Remove the grey gridlines

```{r question6}
ggplot(data = diatask, aes(x = cut, y = price, color = cut)) +
  geom_point(alpha = 0.8, size = 4) +
  labs(
    title = "Diamond Price by Cut",
    x = "Cut",
    y = "Price (US dollars)"
  ) +
  theme_classic()
```

### 7. Add carat and create `diasize`

```{r question7}
diasize <- diamonds[80:100, c("carat", "cut", "price")]

head(diasize)
```

### 8. Plot carat against price

```{r question8}
ggplot(data = diasize, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.8, size = 4) +
  labs(
    title = "Diamond Price by Carat",
    x = "Carat",
    y = "Price (US dollars)",
    color = "Cut"
  ) +
  theme_classic()
```

### 9. Explain why the two small plots are insufficient

```{r question9}
# The plots use only 21 consecutive rows from a dataset containing many
# thousands of diamonds. This small, nonrandom subset does not represent
# the full range of cuts, carat values, or prices. Its prices are also
# concentrated around only a few values. Therefore, the plots are useful
# for practicing R, but they are not sufficient for a reliable general
# conclusion about the relationship between cut and price or between
# carat and price. Because the data are observational, the plots also
# show association rather than proving that one variable causes price.
```

### 10. Construct a stronger plot

```{r question10}
# I used all rows of the diamonds dataset instead of only rows 80 through
# 100. This gives the plot a much broader and more representative range
# of carat and price values.

carat_price <- diamonds[, c("carat", "price")]

nrow(carat_price)

ggplot(data = carat_price, aes(x = carat, y = price)) +
  geom_point(alpha = 0.12, size = 0.7) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    linewidth = 1
  ) +
  labs(
    title = "Diamond Price Increases with Carat",
    subtitle = "All observations in the ggplot2 diamonds dataset",
    x = "Carat",
    y = "Price (US dollars)"
  ) +
  theme_classic()

# I used the complete dataset so the conclusion would not depend on one
# small block of rows. I was not surprised that larger diamonds generally
# had higher prices, but I was surprised by how much prices varied among
# diamonds with similar carat values. This suggests that other variables,
# such as cut, color, and clarity, also help explain price. The plot shows
# a strong positive association, but it does not by itself prove causation.
```

## Notes on `grep()` attempts

1. The pattern must match the values in the data. R is case-sensitive, so `"ideal"` is not the same as `"Ideal"`.

2. Use correct data-frame and function syntax:

```r
dataframe[rows, columns]
function(argument)
variable <- dataframe[1:100, ]
```

3. Tell `grep()` which column to search:

```r
grep("^Good$", diamonds$cut)
```

The `^` and `$` symbols require the whole value to match the word exactly.
