---
title: Penguins! 
format: html
author: Jennifer Danh
---

```{r libSetup}
# Library Setup
library(palmerpenguins)
library(ggplot2)
data(penguins)
```


# Introduction

answer part A questions

# Exploratory Analysis

```{r columnCount}
head(penguins)  
ncol(penguins)   # 8 columns
nrow(penguins)   # 344 penguins
```

Insert note here about columns and what they are
note about how many penguins


```{r 
```

```{r sex}
pmale   <- penguins[penguins$sex == "male", ] 
nrow(pmale)   # 179 male penguins
pfemale <- penguins[penguins$sex == "female", ]
nrow(pfemale) # 176 female penguins
```

note sex counts
equal representation

```{r species}
pAdel <- penguins[penguins$species == "Adelie", ]
pChin <- penguins[penguins$species == "Chinstrap", ]
pGen  <- penguins[penguins$species == "Gentoo", ]
```

talk about species counts here

```{r island}
iBis   <- penguins[penguins$island == "Biscoe", ]
iDream <- penguins[penguins$island == "Dream", ]
iTorg  <- penguins[penguins$island == "Torgersen", ]
```

# Data Visualization

```{r sexWeight}
ggplot(data = penguins, aes(x = sex, y = body_mass_g, fill = sex)) +
  geom_boxplot() +
  labs(title = "Body Mass by Sex", x = "Sex", y = "Body Mass (g)") +
  theme(panel.grid = element_blank())
ggsave("body_mass_sex.png")
```

talk about how body mass showed the most difference in mean between male and female penguins

```{r speciesBill}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(title = "Bill Dimensions by Species", x = "Bill Length (mm)", y = "Bill Depth (mm)")
ggsave("bill_species.png")
```

talk about how bill length scatterplot shows clusters of species

```{r islandPops}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "dodge") +
  labs(title = "Penguin Species by Island", x = "Island", y = "Count")
ggsave("species_island.png")
```

talk about which penguins inhabit which islands