---
title: "Exploring the Palmer Penguins Dataset"
author: "Sofia"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
# Setup chunk for loading required libraries
knitr::opts_chunk$set(echo = TRUE)

# Load the required packages
library(palmerpenguins)  # The Palmer Penguins dataset package
library(tidyverse)       # For data manipulation and visualization
```
## Introduction

The `palmerpenguins` dataset contains size measurements, clutch observations, and blood isotope ratios for three penguin species observed at the Palmer Station Long-Term Ecological Research (LTER) site in Antarctica. Collected by Dr. Kristen Gorman, this data tracks Adélie, Chinstrap, and Gentoo penguins across three islands in the Palmer Archipelago: Biscoe, Dream, and Torgersen. 

The dataset acts as a modern, high-quality alternative to the classic Iris dataset for teaching data science principles. By examining this data, we hope to explore ecological and morphological questions, such as:
* **Species Differentiation:** How reliably can we distinguish between the three penguin species based on their physical traits like bill length, bill depth, and flipper length?
* **Sexual Dimorphism:** Do male and female penguins display significant variations in body mass within or across different species?
* **Islands & Habitats:** Does the geographic distribution across the islands correlate with specific species populations or physical variations?

---

## Exploratory Data Analysis

```{r exploratory-analysis}
# Initial exploration of the 'penguins' dataframe

# Take a quick peek at the structure and first few rows of the data
glimpse(penguins)

# Generate quick summary statistics for all variables
summary(penguins)
```