---
title: "Exploring the Palmer Penguins"
author: "Luis Ponce"
output: html_document
---

```{r setup, include=TRUE}
# Load all libraries needed for this project
library(palmerpenguins)  # provides the 'penguins' dataset
library(dplyr)           # data wrangling (filter, group_by, summarise, etc.)
library(ggplot2)         # data visualization
library(knitr)           # nicer table rendering 
```

## Introduction

The Palmer Penguins dataset was collected and made available by Dr. Kristen
Gorman and the Palmer Station Long Term Ecological Research (LTER) Network,
part of the US Long Term Ecological Research Network based in Antarctica.
The data include physical measurements and other observations for three
penguin species — Adelie, Chinstrap, and Gentoo — recorded at three islands
in the Palmer Archipelago near Antarctica: Biscoe, Dream, and Torgersen.

This dataset was originally used to examine how body size and shape vary across penguin species, sexes, and
populations, and how these physical traits relate to ecological factors
such as dietary niche. The original 2014 study (https://doi.org/10.1371/journal.pone.0090081) used measurements like flipper
length, body mass, and bill dimensions, along with blood isotope data, to
explore differences between species and sexes and to understand foraging
ecology in these penguins.

The objective is to use the `penguins` dataset to answer some
introductory questions:

-How large is this dataset, and how is it structured?
* Are males and females equally represented in the data?
* Are all three species and all three islands equally represented?
* Which physical measurements best distinguish male penguins from female
  penguins?
* Which physical measurements best distinguish the three species from one
  another?
* Which islands are home to which species?

## Exploratory Analysis

```{r explore-setup}
# Explores the structure of the dataset
str(penguins)
```

### How many data columns are in `penguins`?

```{r num-columns}
# ncol() gives the number of columns (variables) in the dataset. 
ncol(penguins)
```

The dataset contains several columns, including species, island, bill
length and depth, flipper length, body mass, sex, and year of collection.

### How many penguins do we have data for?
344; 152 Adelie, 124 Gentoo, and 68 Chinstrap.
```{r num-rows}
# nrow() gives the number of rows of individual penguin observations
nrow(penguins)
```

### Do we have equal representation of males and females?

```{r sex-breakdown}
# table() with useNA shows counts per sex, including any missing (NA) values
# This is important because some penguins have NA for sex - a straightforward
# grep() or == comparison could silently drop or mishandle these rows
table(penguins$sex, useNA = "ifany")
```

```{r male-female-split}
# Create a dataframe with only male penguins
# NOTE: filter() with == "male" will naturally exclude NA sex values,
# which is what we want - we don't want NAs miscategorized as male
pmale <- penguins %>% filter(sex == "male")

# Create a dataframe with only female penguins
pfemale <- penguins %>% filter(sex == "female")

# Double check the counts add up sensibly (they won't sum to nrow(penguins)
# because of the NA sex entries we intentionally excluded from both)
nrow(pmale)
nrow(pfemale)
nrow(penguins) - nrow(pmale) - nrow(pfemale)  # number of penguins with unknown sex
```

Males and females are fairly evenly represented overall, but there are a
small number of penguins with missing sex data. Those individuals were
purposely excluded from both `pmale` and `pfemale` rather than being
incorrectly categorized into one group.

### Do we have equal representation of each species?

```{r species-breakdown}
# Count how many observations belong to each species
table(penguins$species)
```

```{r species-split}
# Create separate dataframes for each species
pAdel <- penguins %>% filter(species == "Adelie")
pChin <- penguins %>% filter(species == "Chinstrap")
pGen  <- penguins %>% filter(species == "Gentoo")

nrow(pAdel)
nrow(pChin)
nrow(pGen)
```

Species representation is not perfectly balanced: Adelie penguins(152) have the
most observations, while Chinstrap penguins (68) have the fewest.

### Do we have equal representation of each island?

```{r island-breakdown}
# Count how many observations come from each island
table(penguins$island)
```

```{r island-split}
# Create separate dataframes for each island
iBis   <- penguins %>% filter(island == "Biscoe")
iDream <- penguins %>% filter(island == "Dream")
iTorg  <- penguins %>% filter(island == "Torgersen")

nrow(iBis)
nrow(iDream)
nrow(iTorg)
```

Island representation is also uneven: Biscoe has the most observations,
while Torgersen has the fewest. This makes sense biologically, since not
every species was observed on every island.

## Data Visualization

### What distinguishes male and female penguins?

```{r sex-boxplot}
# Boxplot comparing body mass between sexes, faceted by species
# so we aren't confounding species differences with sex differences
penguins %>%
  filter(!is.na(sex)) %>%
  ggplot(aes(x = sex, y = body_mass_g, fill = sex)) +
  geom_boxplot(alpha = 0.8) +
  facet_wrap(~ species) +
  scale_fill_manual(values = c("female" = "#E69F00", "male" = "#56B4E9")) +
  labs(
    title = "Body Mass by Sex, Within Each Species",
    x = "Sex",
    y = "Body Mass (g)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

First, body mass; there was a clear difference between males and females. 
This was observable across all 3 species of penguines. Males were in most cases heavier than the females.

### What distinguishes the three species?

```{r species-scatter}
# Scatterplot of bill length vs. bill depth, colored by species
# This combination is known to strongly separate the three species
penguins %>%
  filter(!is.na(bill_length_mm)) %>%
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species, shape = species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("Adelie" = "#E69F00", "Chinstrap" = "#009E73", "Gentoo" = "#0072B2")) +
  labs(
    title = "Bill Length vs. Bill Depth by Species",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)",
    color = "Species",
    shape = "Species"
  ) +
  theme_minimal(base_size = 13)
```

Their bills dimensions is accurate way to distinguish between the three species. 
Adelie have short and deep bills, Gentoo have long and shallow bills, and Chinstrap 
longer and deep bills. 
### Which species inhabit each island?

```{r island-species-bar}
# Bar chart showing counts of each species observed on each island
penguins %>%
  ggplot(aes(x = island, fill = species)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("Adelie" = "#E69F00", "Chinstrap" = "#009E73", "Gentoo" = "#0072B2")) +
  labs(
    title = "Penguin Species Counts by Island",
    x = "Island",
    y = "Number of Penguins",
    fill = "Species"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 0))
```

Adelie live exclusively in Torgersen Island, but some of them live in Dream Island. Dream island is
shared between the Chinstrap and Adelie populations. Biscoe Island is also shared by Adelie
and Gentoo. No island is exclusive to one population. 



:::
