---
title: "Exploring the Palmer Penguins Dataset"
author: "Sofia"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
# Setup chunk for loading required libraries
knitr::opts_chunk$set(echo = TRUE)

# Load the required packages
library(palmerpenguins)  # The Palmer Penguins dataset package
library(tidyverse)       # For data manipulation and visualization
```The palmerpenguins dataset contains size measurements, clutch observations, and blood isotope ratios for three penguin species observed at the Palmer Station Long-Term Ecological Research (LTER) site in Antarctica. Collected by Dr. Kristen Gorman, this data tracks Adélie, Chinstrap, and Gentoo penguins across three islands in the Palmer Archipelago: Biscoe, Dream, and Torgersen.

By examining this data, we hope to explore ecological and morphological questions, such as how reliably we can distinguish between the three penguin species based on physical traits, whether they display sexual dimorphism, and how populations differ across islands.
# Count the number of columns (variables) in the dataset
ncol(penguins)

# Count the number of rows (individual penguins) in the dataset
nrow(penguins)

# Count how many penguins fall into each sex category (including missing data)
table(penguins$sex, useNA = "ifany")

# Create a dataframe containing only male penguins using exact matching
pmale <- penguins %>% 
  filter(sex == "male")

# Create a dataframe containing only female penguins using exact matching
pfemale <- penguins %>% 
  filter(sex == "female")

# Count the number of observations for each penguin species
table(penguins$species)

# Subset the data to isolate Adélie penguins
pAdel <- penguins %>% 
  filter(species == "Adelie")

# Subset the data to isolate Chinstrap penguins
pChin <- penguins %>% 
  filter(species == "Chinstrap")

# Subset the data to isolate Gentoo penguins
pGen <- penguins %>% 
  filter(species == "Gentoo")

# Count how many penguins were observed on each island
table(penguins$island)

# Subset the data to isolate penguins from Biscoe Island
iBis <- penguins %>% 
  filter(island == "Biscoe")

# Subset the data to isolate penguins from Dream Island
iDream <- penguins %>% 
  filter(island == "Dream")

# Subset the data to isolate penguins from Torgersen Island
iTorg <- penguins %>% 
  filter(island == "Torgersen")
```
---

## Data Visualization

```{r data-visualization, warning=FALSE, message=FALSE}
# 1. Distinguishing Male vs. Female Penguins (Boxplot)
# Bill depth and body mass tend to show clear separation between sexes.
ggplot(data = drop_na(penguins, sex), aes(x = sex, y = body_mass_g, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Body Mass Distribution by Sex",
    x = "Sex",
    y = "Body Mass (g)"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")

# 2. Distinguishing Penguin Species (Scatter Plot)
# Plotting bill length vs. flipper length cleanly separates the species.
ggplot(data = penguins, aes(x = bill_length_mm, y = flipper_length_mm, color = species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title = "Species Differentiation by Bill and Flipper Length",
    x = "Bill Length (mm)",
    y = "Flipper Length (mm)",
    color = "Species"
  ) +
  theme_light() +
  scale_color_viridis_d(option = "plasma")

# 3. Species Distribution Across Islands (Bar Plot)
# Shows exactly which species inhabit which island sites.
ggplot(data = penguins, aes(x = island, fill = species)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Species Counts Across Antarctic Island Sites",
    x = "Island",
    y = "Number of Penguins",
    fill = "Species"
  ) +
  theme_classic() +
  scale_fill_dark()
```
