---
title: "Computing Exercise 9: Palmer Penguins Analysis"
author: "Sophie Earp"
output: html_document
---
```{r setup, include=TRUE}
# Load required libraries
library(palmerpenguins)
library(ggplot2)
library(dplyr)
```
## Introduction

The **Palmer Penguins** dataset contains size measurements, species classifications, and island locations for adult penguins observed near Palmer Station, Antarctica. The data were collected by Dr. Kristen Gorman and the Palmer Station Long Term Ecological Research (LTER) program. 

In this analysis, we aim to explore:
* How penguin populations are distributed across different species and island locations.
* Differences in representation between male and female penguins.
* Key physical features (such as flipper length, bill dimensions, and body mass) that best distinguish species and sexes.
```{r exploratory-analysis}
# 1. Check total number of columns in the dataset
ncol(penguins)

# 2. Check total number of penguins (rows) in the dataset
nrow(penguins)

# 3. Separate by sex to test equal representation
# Filtering male penguins
pmale <- penguins[penguins$sex == "male" & !is.na(penguins$sex), ]

# Filtering female penguins (avoiding grep due to partial matching edge cases)
pfemale <- penguins[penguins$sex == "female" & !is.na(penguins$sex), ]

# 4. Separate into new dataframes by species
pAdel <- penguins[penguins$species == "Adelie" & !is.na(penguins$species), ]
pChin <- penguins[penguins$species == "Chinstrap" & !is.na(penguins$species), ]
pGen  <- penguins[penguins$species == "Gentoo" & !is.na(penguins$species), ]

# 5. Separate into new dataframes by island
iBis  <- penguins[penguins$island == "Biscoe" & !is.na(penguins$island), ]
iDream <- penguins[penguins$island == "Dream" & !is.na(penguins$island), ]
iTorg  <- penguins[penguins$island == "Torgersen" & !is.na(penguins$island), ]
```

### Summary of Findings:
* **Dimensions:** The `penguins` dataset contains 8 columns and 344 rows (individual penguins).
* **Sex Representation:** There are 168 male penguins and 165 female penguins recorded (with 11 missing values for sex). Representation between sexes is nearly equal.
* **Species Representation:** The dataset is dominated by Adelie penguins (152), followed by Gentoo (124) and Chinstrap (68).
* **Island Distribution:** Biscoe Island has 168 penguins, Dream Island has 124, and Torgersen Island has 52.

```{r data-visualizations}
# Plot 1: Boxplot showing body mass differences between males and females across species
ggplot(data = na.omit(penguins), aes(x = species, y = body_mass_g, fill = sex)) +
  geom_boxplot(alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = c("female" = "darkorange", "male" = "cyan4")) +
  labs(
    title = "Body Mass Distribution by Species and Sex",
    x = "Penguin Species",
    y = "Body Mass (g)",
    fill = "Sex"
  )

# Plot 2: Scatter plot of Flipper Length vs. Bill Length distinguishing species
ggplot(data = na.omit(penguins), aes(x = bill_length_mm, y = flipper_length_mm, color = species, shape = species)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_classic() +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Flipper Length vs. Bill Length by Species",
    x = "Bill Length (mm)",
    y = "Flipper Length (mm)",
    color = "Species",
    shape = "Species"
  )

# Plot 3: Bar plot showing species distribution across Antarctic Islands
ggplot(data = penguins, aes(x = island, fill = species)) +
  geom_bar(position = "dodge") +
  theme_light() +
  scale_fill_viridis_d(option = "plasma") +
  labs(
    title = "Penguin Species Distribution Across Islands",
    x = "Island",
    y = "Count",
    fill = "Species"
  )
```
