---
title: "Exploring the Palmer Penguins Dataset"
author: "Lexi Love"
output: html_document
---

```{r setup, include=FALSE}
library(palmerpenguins)
library(ggplot2)
```

# Introduction
The Palmer penguins datasets contains measurements collected from Adelie, Chinstrap, and Gentoo penguins living in Biscoe, Dream, and Torgersen islands near Palmer station in Antartica. 
Researchers collected information about each penguin, including its species, sex, bill measurement, flipper length and body mass.
I will explore the palmer penguins dataset to compare male and female penguins, examine differences between the three species, and determine which species are found on each island. 
I also hope to identify patterns in the penguins physical characterstics by creating tables and graphs. 

# Exploratory Analysis

This section explores the Palmer Penguins dataset by examining the number of variables, the number of penguins, and how the data are distributed by sex, species, and island.

## Number of Columns

```{r}
# Count the number of columns
ncol(penguins)
```
## Number of Penguins

Next, I will determine how many penguins are included in the dataset.

```{r}
# Count the number of penguins
nrow(penguins)
```
## Male vs. Female Representation

This code checks whether there are equal numbers of male and female penguins.

```{r}
# Count males and females
table(penguins$sex)
```
## Male Penguin Dataframe

This creates a new dataframe containing only male penguins.

```{r}
# Create dataframe of male penguins
pmale <- penguins[which(penguins$sex == "male"), ]
```

## Female Penguin Dataframe

This creates a new dataframe containing only female penguins.

```{r}
# Create dataframe of female penguins
pfemale <- penguins[which(penguins$sex == "female"), ]
```

## Do we have equal representation of each Species?

This code counts the number of penguins in each species. 

```{r}
# Count the number of penguins in each species
table(penguins$species)
```

## Create a dataframe of Adelie penguins

This dataframe contains only Adelie penguins.

```{r}
# Create a dataframe containing only Adelie penguins
pAdel <- penguins[penguins$species == "Adelie", ]
```

## Create a dataframe of Chinstrap penguins

This dataframe contains only Chinstrap penguins.

```{r}
# Create a dataframe containing only Chinstrap penguins
pChin <- penguins[penguins$species == "Chinstrap", ]
```

## Create a dataframe of Gentoo penguins

This dataframe contains only Gentoo penguins.

```{r}
# Create a dataframe containing only Gentoo penguins
pGen <- penguins[penguins$species == "Gentoo", ]
```

## Do we have equal representation of each island?

Finally, I will determine how many penguins are found on each island.

```{r}
# Count the number of penguins on each island
table(penguins$island)
```

## Create a dataframe of Biscoe Island penguins

This dataframe contains only penguins from Biscoe Island.

```{r}
# Create a dataframe containing only Biscoe Island penguins
iBis <- penguins[penguins$island == "Biscoe", ]
```

## Create a dataframe of Dream Island penguins

This dataframe contains only penguins from Dream Island.

```{r}
# Create a dataframe containing only Dream Island penguins
iDream <- penguins[penguins$island == "Dream", ]
```

## Create a dataframe of Torgersen Island penguins

This dataframe contains only penguins from Torgersen Island.

```{r}
# Create a dataframe containing only Torgersen Island penguins
iTorg <- penguins[penguins$island == "Torgersen", ]
```

# Data Visualization

This section will use graphs to explore differences between penguin sex, species, and island location.

## Male vs. Female Penguins

The scatterplot below compares flipper length and body mass for male and female penguins. These measurements help show how the two groups differ.

```{r}
#Create a scatter plot comparing male and female penguins
ggplot(data = penguins[!is.na(penguins$sex), ], aes(x = flipper_length_mm, y = body_mass_g, color = sex))+
    geom_point(size = 3, alpha = 0.8) + labs(title = "Male vs. Female Penguins", 
    x = "Flipper Length (mm)", y = "Body Mass (g)", color = "Sex" ) +
    theme_minimal() + theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```

This graph shows that male penguins generally have longer flippers and greater body mass than female penguins, although there is some overlap.


## Penguin Species Comparison

The boxplot below compares bill length among the three penguin species and I chose bill length because its one of the measurements that helps distinguish the species.

```{r}
# Create a boxplot comparing bill length by species
ggplot(data = penguins, aes(x = species, y = bill_length_mm, fill = species)) + 
    geom_boxplot() + labs(title = "Bill Length by Penguin Species", x = "Species", y = "Bill Length (mm)")+
    theme_minimal() + theme(plot.title = element_text(face = "bold"), legend.position = "none")
```

This graph shows that bill length differs among the three penguin species, making it a useful measurement for distinguishing them.

## Penguin Species on Each Island 

The bar chart below shows which penguin species are found on each Antarctic island.

```{r}
# Create a bar chart showing penguin species on each island
ggplot(data = penguins, aes(x = island, fill = species)) + 
    geom_bar(position = "dodge") + labs(title = "Penguin Species on Each Island", x = "Island", y = "Number of Penguins", fill = "Species")+
    theme_minimal() + theme(plot.title = element_text(face = "bold"), legend.position = "right")
```

This graph shows that Adelie penguins are found on all three islands, Chinstrap penguins are found only on Dream Island,
and Gentoo penguins are found only on Biscoe Island. 