---
title: "Computing Exercise 9: Penguins in R"
author: "Devin Gay"
output: rmarkdown::github_document
---

## Penguins innit
Source for the output in the YAML is from [How can i see output of .rmd in github?](https://stackoverflow.com/questions/39814916/how-can-i-see-output-of-rmd-in-github)

``` {r libraries, echo = TRUE, message=FALSE}
library(rmarkdown)
library(ggplot2)
library(palmerpenguins)
library(knitr)
```

`palmerpenguins` is derived from data collected on three species of Antarctic penguins from three islands. The study was conducted in the interest of collecting sex-specific foraging data. It was predicted that all three species would exhibit sexually dimorphic feeding, only aligning when the foraging habitat was particularly favorable.
``` {r penguinhelp, echo=F}
kable(head(penguins))
```

1. How many data columns are in 'penguins'?

There are eight columns. dim() gives the results of nrow() and ncol().

The columns represent variables, which in this case are species, island, bill length, bill depth, flipper length, body mass, sex, and the year of observation
``` {r columns}
ncol(penguins)

dim(penguins)
```

2. How many penguins do we have data for?

There is data for 344 penguins
``` {r rows}
nrow(penguins)
```

3. Do we have equal representations of males and females in the dataset? 

There are 168 males and 165 females, this is roughly equal. The sex of 11 penguins is unknown.

Dataframes based on sex:
``` {r representation}
length(which(penguins$sex=="male")) 
length(which(penguins$sex=="female")) 
sum(is.na(penguins$sex))
```

``` {r dfsex}
pmale <- penguins[grep("^male",penguins$sex),]
nrow(pmale)
pfemale <- penguins[grep("^female",penguins$sex),]
nrow(pfemale)

kable(head(pmale))
kable(head(pfemale))
```


4. Do we have equal representation of each species in the dataset? 

There are 152 Adelie, 68 Chinstrap, and 124 Gentoo penguins. There is not equal representation of species

Dataframes based on species:

``` {r dfspp}
unique(penguins$species) # Adelie Chinstrap Gentoo
pAdel <- penguins[grep("Adelie",penguins$species),]
nrow(pAdel) # 152
pChin <- penguins[grep("Chinstrap",penguins$species),]
nrow(pChin) # 68
pGen <- penguins[grep("Gentoo",penguins$species),]
nrow(pGen) # 124

152+68+124 # = 344
```

5. Do we have equal representation of each island in the dataset?

There are 168 penguins from Biscoe island, 124 from Dream island, and 52 from Torgersen island. There is not equal representation of each island.

``` {r islands}
unique(penguins$island) # Biscoe Dream Torgersen
iBis <- penguins[grep("Biscoe",penguins$island),]
nrow(iBis) # 168
iDream <- penguins[grep("Dream",penguins$island),]
nrow(iDream) # 124
iTorg <- penguins[grep("Torgersen",penguins$island),]
nrow(iTorg) # 52

168+124+52
```

## Visualization with ggplot2

1a. What measured values in the dataset most clearly distinguish male and female penguins?

I chose body mass as a measure of sexual dimorphism. Body mass is often a sexually dimorphic trait, and this remains largely true with these penguins. Body mass was found to me one of the most sexually dimorphic traits among Adelie and Gentoo penguins, but was also the least dimorphic in Chinstraps. No trait was consistently dimorphic among all three species, with body mass, bill length, and bill depth all being good predictors for two of the three species, and relatively poor for the third.
``` {r plotsex}
ggplot(data=subset(penguins, !is.na(penguins$sex)),aes(x=sex,y=body_mass_g)) +
geom_boxplot(data=subset(penguins, !is.na(penguins$sex)),aes(color=sex,fill=sex),alpha=.6) +
stat_summary(fun=mean,geom="point") +
theme_classic() +
scale_x_discrete(labels=c("Female","Male")) +
labs(y="Body Mass (g)",x="") +
theme(legend.position="none") +
facet_wrap(~species)
```

1b. What measured values in the dataset most clearly distinguish each of the species of penguins?

Because this is a study on feeding ecology, I chose bill length and depth as species predictors. From the plot there is some decent species-based clustering. Adelie penguins tend to have deeper, shorter bills. Gentoos tend to have longer, shallower bills. Chinstraps tend to have longer and deeper bills than the other two species.

``` {r plotspp, warning = FALSE,message=F}
ggplot(penguins,aes(x=bill_depth_mm,y=bill_length_mm)) +
geom_point(penguins,mapping=aes(color=species)) +
geom_smooth(method="lm",mapping=aes(color=species),se=F) +
theme_classic() +
labs(x="Bill depth (mm)",y="Bill length (mm)")
```

1c. Which penguins species inhabit each of the Antarctic island sites? 

I've never used geom_jitter() like this before, I think it's fun.

A more familiar option is using geom_bar. position="fill" creates stacked bars to show proportions.

Adelie penguins are found on all three islands and are the only species recorded on Torgersen. Gentoos are only found on Biscoe and Chinstraps are only found on Dream island. I think the bar plot is more effective here as it effectively shows the species proportions on each island. Gentoo and Chinstrap penguins dominate on their respective islands despite the presence of Adelie on all three.
``` {r islandplot}
ggplot(penguins,aes(x=island,y=species)) +
geom_jitter(aes(color=species),alpha=0.6) +
labs(x="Island",y="Species") +
theme_classic()

ggplot(penguins,aes(x=island,fill=species)) +
geom_bar(position="fill") +
theme_classic() +
scale_y_discrete(expand=c(0,0)) +
labs(x="Island",y="")
```