---
title: "Exploring the Palmer Penguins Dataset"
author: "Sidney"
output: html_document
---
```

```{r setup}
# Loading the packages I need for this project.
# palmerpenguins gives me the penguins dataset.
# tidyverse helps me work with data and make graphs.
library(palmerpenguins)
library(tidyverse)
```

# Introduction

The Palmer Penguins dataset contains measurements of three penguin species from islands near Palmer Station in Antarctica. The data includes things like bill length, bill depth, flipper length, and body mass. I want to explore questions like how many penguins there are, how males and females differ, and how the species compare.

```{r}
# Looking at the penguins dataset to make sure it loaded.
penguins
```

# Exploratory Analysis

```{r}
# Number of columns in the dataset.
ncol(penguins)
```

```{r}
# Number of rows (penguins) in the dataset.
nrow(penguins)
```

```{r}
# Checking how many males and females there are.
table(penguins$sex)
```

```{r}
# Making a smaller dataset with only male penguins.
pmale <- penguins %>% filter(sex == "male")
pmale
```

```{r}
# Making a smaller dataset with only female penguins.
pfemale <- penguins %>% filter(sex == "female")
pfemale
```

```{r}
# Checking how many penguins belong to each species.
table(penguins$species)
```

```{r}
# Making separate datasets for each species.
pAdel <- penguins %>% filter(species == "Adelie")
pChin <- penguins %>% filter(species == "Chinstrap")
pGen  <- penguins %>% filter(species == "Gentoo")
```

```{r}
# Checking how many penguins come from each island.
table(penguins$island)
```

```{r}
# Making separate datasets for each island.
iBis   <- penguins %>% filter(island == "Biscoe")
iDream <- penguins %>% filter(island == "Dream")
iTorg  <- penguins %>% filter(island == "Torgersen")
```

# Data Visualization

## Male vs Female Differences
This plot compares male and female penguins to see which measurements differ the most.

```{r}
ggplot(penguins, aes(x = sex, y = body_mass_g, fill = sex)) +
  geom_boxplot() +
  labs(title = "Body Mass Differences Between Male and Female Penguins",
       x = "Sex",
       y = "Body Mass (g)") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")
```

## Species Differences
This plot shows how flipper length varies between Adelie, Chinstrap, and Gentoo penguins.

```{r}
ggplot(penguins, aes(x = species, y = flipper_length_mm, fill = species)) +
  geom_violin() +
  labs(title = "Flipper Length Differences Between Penguin Species",
       x = "Species",
       y = "Flipper Length (mm)") +
  theme_light() +
  scale_fill_brewer(palette = "Dark2")
```

## Species on Each Island
This plot shows which penguin species live on each island.

```{r}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "dodge") +
  labs(title = "Penguin Species on Each Island",
       x = "Island",
       y = "Number of Penguins") +
  theme_bw() +
  scale_fill_brewer(palette = "Set1")
```