---
title: Startup in R with Project Quarto! 
format: html
---
## Title 

The purpose of a Markdown file is to keep explanations for computational problems and the code we use to solve them next to each other in the same document. This is where you can write description about your project and the data you are working with. 

How many frames do you see in this space and what you they do? 

There are two frames. One prints hello world. the other installs packages. 

Next up: Let's make a space for writing code that R can understand: 

```{r}
install.packages("rmarkdown") 


print("Hello, World! im github!")
```

## Instructions

We are outside of the code because the code chunk starts with ```{r} and ends with the same ``` mark. Only things inbetween the three accent marks (they are calle 'grave' marks) is seen to be code by the computer. 

NOTE: The grave accent is the key at the top left of the keyboard, NOT an apostrophe. It also has the tilde (~) mark on it. 


```{r second}
#We started another piece of code! Look how we can go back and forth between writing explainations and writing code! 
print("IDK WHAT THIS STUFF EVEN MEANS")
summary(rnorm(100)) 
```
